Long--term perspectives of social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Session "Health and Mortality in Arctic Populations" at The Ninth International Congress on Arctic Social Sciences (ICASS IX) June 10, 2017, Umeå}

### 2017-06-08 Umea.Rnw
Small revisions

\item[sex]
\item[birth date] By quarter (15 Feb, May, Aug, Nov)
\item[death date] By day (\emp{Socialstyrelsen})
\item[cause of death] from \emp{Socialstyrelsen}
\item[parish] Urban/Rural
\item[Social class] elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005, three categories.
\item[period]  1851--1875, \ldots,  2008--2013 (10 periods).
%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

\begin{itemize}
\item \emp{Improved} survival characterised all {\sc Hisclass}es.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
\item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection between income inequality and social differences in mortality.
\end{itemize}

@@

\begin{itemize}
\item The development of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
\item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).
\item The role of \emp{living standard} for mortality decline (McKeown).
\item \emp{Gender} and mortality (Watkins, Mckinnon, Janssens).
\item Social conditions as \emp{fundamental causes of death} (Link/Phelan).

>>>>

\item[sex]
\item[birth date] By quarter (15 Feb, May, Aug, Nov)
\item[death date] By day (\emp{Socialstyrelsen})
\item[cause of death] from \emp{Socialstyrelsen}
\item[parish] Urban/Rural
\item[Social class] Elite, middle, worker. Household definition 1851--1950. Individual definition 1960--2013
\item[income] 1990--2005, grouped by quartiles.
\item[education] 1990--2005, three categories.
\item[period]  1851--1875, \ldots,  2008--2013 (10 periods). No results presented for women 1961-70, 1971-80 due to missing information on class.
%\item[cohort] 1761--1771, \ldots, 1971--1974 (22 birth cohorts).

@@

\begin{itemize}
\item \emp{Improved} survival characterised all social classes.
\item \emp{Late} appearance of the “modern” pattern of social differences in mortality among men
\item Distinct \emp{gender} differences in the development.
\item The \emp{upper} class(es) gradually take over as the \emp{most healthy} group.
\item No \emp{qualitative} but partly \emp{quantitative} differences between those in working age compared to the retired population.
\item \emp{Educational} and \emp{income} differences during the last decades conform to what we found for class differences.
\item In the long run, no clear connection between income inequality and social differences in mortality.
\end{itemize}

@@

\begin{itemize}
\item The \emp{development} of social health inequalities and the role of \emp{health-related behavior} in different phases of the transition (Antonovsky and Omran).
\item Inequality as a \emp{contextual} health risk (Wilkinson/Pickett).
\item \emp{Status syndrome} (Marmot)
\item The role of \emp{living standard} for mortality decline (McKeown).
\item \emp{Gender} and mortality (Watkins, Mckinnon, Janssens).
\item Social conditions as \emp{fundamental causes of death} (Link/Phelan).
